---
title: "Modern Data Mining - HW 4"
author:
- Graham Branscom
- Mahika Calyanakoti
- Andrew Raine
date: 'Due: 11:59Pm,  April 7th, 2024'
output:
  pdf_document:
    toc: yes
    toc_depth: '4'
  html_document:
    code_folding: show
    highlight: haddock
    theme: lumen
    toc: yes
    toc_depth: 4
    toc_float: yes
  word_document:
    toc: yes
    toc_depth: '4'
urlcolor: blue  
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.height=4, fig.width=6, warning = F)
if (!require("pacman")) install.packages("pacman")
pacman::p_load(randomForest, tree, ISLR, rpart, rattle, pROC, partykit, ggplot2, glmnet, leaps, dplyr, keras, neuralnet, imager, ranger, xgboost)
```

test


# Overview

In this homework, we will explore the transition from linear models to more flexible, tree-based methods and ensemble techniques in predictive modeling. Unlike linear models, a model-free approach, such as binary decision trees, offers a more intuitive understanding by illustrating direct relationships between predictors and responses. Although simple, binary decision trees are highly interpretable and can unveil valuable insights.

However, to harness greater predictive power, we can extend beyond a single decision tree. By aggregating multiple models, particularly those that are uncorrelated, we significantly enhance our predictive accuracy. A prime example of this concept is the RandomForest algorithm. Here, we create a multitude of decision trees through bootstrap sampling – a method where each tree is built from a random subset of data and variables. The aggregation of these diverse trees results in a robust final prediction model.

Ensemble methods extend this idea further by combining various models to improve predictive performance. This could involve averaging or taking a weighted average of numerous distinct models. Often, this approach surpasses the predictive capability of any individual model at hand, providing a powerful tool for tackling complex data mining challenges.

Boosting, particularly Gradient Boosting Machines, stands out as another potent predictive method. Unlike traditional ensemble techniques that build models independently, boosting focuses on sequentially improving the prediction by specifically targeting the errors of previous models. Each new model incrementally reduces the errors, leading to a highly accurate combined prediction. 

All the methods mentioned above  can handle diverse types of data and predict outcomes ranging from continuous to categorical responses, including multi-level categories.

In Homework 4, we will delve into these advanced techniques, moving beyond the limitations of linear models and exploring the expansive potential of trees, ensembles, and boosting in modern data mining. This journey will provide you with a solid foundation in leveraging sophisticated algorithms to uncover deeper insights and achieve superior predictive performance. 


## Objectives


- Understand trees
    + single tree/displaying/pruning a tree
    + RandomForest
    + Ensemble idea
    + Boosting 

- R functions/Packages
    + `tree`, `RandomForest`, `ranger`
    + Boosting functions
    
- Json data format

- text mining
    + bag of words
  

Data needed:

+ `IQ.Full.csv`
+ `yelp_review_20k.json`

# Problem 1: IQ and successes

## Background: Measurement of Intelligence 

Case Study:  how intelligence relates to one's future successes?

**Data needed: `IQ.Full.csv`**

ASVAB (Armed Services Vocational Aptitude Battery) tests have been used as a screening test for those who want to join the army or other jobs. 

Our data set IQ.csv is a subset of individuals from the 1979 National Longitudinal Study of 
Youth (NLSY79) survey who were re-interviewed in 2006. Information about family, personal demographic such as gender, race and education level, plus a set of ASVAB (Armed Services Vocational Aptitude Battery) test scores are available. It is STILL used as a screening test for those who want to join the army! ASVAB scores were 1981 and income was 2005. 

**Our goals:** 

+ Is IQ related to one's successes measured by Income?
+ Is there evidence to show that Females are under-paid?
+ What are the best possible prediction models to predict future income? 


**The ASVAB has the following components:**

+ Science, Arith (Arithmetic reasoning), Word (Word knowledge), Parag (Paragraph comprehension), Numer (Numerical operation), Coding (Coding speed), Auto (Automative and Shop information), Math (Math knowledge), Mechanic (Mechanic Comprehension) and Elec (Electronic information).
+ AFQT (Armed Forces Qualifying Test) is a combination of Word, Parag, Math and Arith.
+ Note: Service Branch requirement: Army 31, Navy 35, Marines 31, Air Force 36, and Coast Guard 45,(out of 100 which is the max!) 

**The detailed variable definitions:**

Personal Demographic Variables: 

 * Race: 1 = Hispanic, 2 = Black, 3 = Not Hispanic or Black
 * Gender: a factor with levels "female" and "male"
 * Educ: years of education completed by 2006
 
Household Environment: 
 
* Imagazine: a variable taking on the value 1 if anyone in the respondent’s household regularly read
	magazines in 1979, otherwise 0
* Inewspaper: a variable taking on the value 1 if anyone in the respondent’s household regularly read
	newspapers in 1979, otherwise 0
* Ilibrary: a variable taking on the value 1 if anyone in the respondent’s household had a library card
	in 1979, otherwise 0
* MotherEd: mother’s years of education
* FatherEd: father’s years of education

Variables Related to ASVAB test Scores in 1981 (Proxy of IQ's)

* AFQT: percentile score on the AFQT intelligence test in 1981 
* Coding: score on the Coding Speed test in 1981
* Auto: score on the Automotive and Shop test in 1981
* Mechanic: score on the Mechanic test in 1981
* Elec: score on the Electronics Information test in 1981

* Science: score on the General Science test in 1981
* Math: score on the Math test in 1981
* Arith: score on the Arithmetic Reasoning test in 1981
* Word: score on the Word Knowledge Test in 1981
* Parag: score on the Paragraph Comprehension test in 1981
* Numer: score on the Numerical Operations test in 1981

Variable Related to Life Success in 2006

* Income2005: total annual income from wages and salary in 2005. We will use a natural log transformation over the income.


**Note: All the Esteem scores shouldn't be used as predictors to predict income**

## 1. EDA: Some cleaning work is needed to organize the data. 

+ The first variable is the label for each person. Take that out.
+ Set categorical variables as factors. 
+ Make log transformation for Income and take the original Income out
+ Take the last person out of the dataset and label it as **Michelle**. 
+ When needed, split data to three portions: training, testing and validation (70%/20%/10%)
  - training data: get a fit
  - testing data: find the best tuning parameters/best models
  - validation data: only used in your final model to report the accuracy.
  
We remove the first col (subject id) and set some columns as factors. These are the remaining features:
```{r, echo=F, results=F}
# load in data
iq_df <- read.csv("./data/IQ.Full.csv")

# remove first col
iq_df <- iq_df[,-1]
```
```{r, echo=F, results=T}
# set categorical variables as factors
iq_df$Imagazine <- as.factor(iq_df$Imagazine)
iq_df$Inewspaper <- as.factor(iq_df$Inewspaper)
iq_df$Ilibrary <- as.factor(iq_df$Ilibrary)
iq_df$Race <- as.factor(iq_df$Race)
iq_df$Gender <- as.factor(iq_df$Gender)
```

```{r, echo=F, results=T}
# log transform income and remove original income
iq_df$log_income <- log(iq_df$Income2005)
iq_df <- subset(iq_df, select = -Income2005)
str(iq_df)
```

We get the last row as **Michelle** and then remove it.
```{r, echo=F, results=T}
michelle <- iq_df[2584,]
iq_df <- iq_df[-2584,]
# dim(iq_df)
michelle
```
We split the data into training (70%), testing (20%), and validation (10%) sets. Then print out the dimensions of the train, test, and validation dataframes.
```{r, echo=F, results=T}
set.seed(123)

n <- nrow(iq_df)
train.index <- sample(n, n*0.70)
data.train <- iq_df[train.index,]
data.temp <- iq_df[-train.index, ]
n_temp <- nrow(data.temp)
test.index <- sample(n_temp, n_temp*2/3)
data.test <- data.temp[test.index,]
data.val <- data.temp[-test.index,]

dim(data.train)
dim(data.test)
dim(data.val)
```


## 2. Factors affect Income

We start with linear models to answer the questions below.

i. To summarize ASVAB test scores, create PC1 and PC2 of 10 scores of ASVAB tests and label them as
ASVAB_PC1 and ASVAB_PC2. Give a quick interpretation of each ASVAB_PC1 and ASVAB_PC2 in terms of the original 10 tests.

PC1 is a weighted sum of all of the 10 scores. All of PC1's coefficients are around 0.30, with some variation. PC2 is proportional to the sum of all scores except science, auto, mechanic, and elec, which are weighted negatively. See the coefficients below:
```{r, echo=F, results=T}
# ASVAB columns
asvab_cols <- c("Science", "Arith", "Word", "Parag", "Numer", "Coding", "Auto", "Math", "Mechanic", "Elec")

# ASVAB intermediate df
asvab_df <- iq_df[, asvab_cols]

# find asvab pc1
pca <- prcomp(asvab_df, scale=TRUE, center=T)
pc1_2 <- pca$rotation[,c(1,2)]
pc1_2
```


ii. Is there any evidence showing ASVAB test scores in terms of ASVAB_PC1 and ASVAB_PC2, might affect the Income?  Show your work here. You may control a few other variables, including gender.

We added the PC1 and PC2 scores for each row in `iq_df`.
```{r, echo=F, results=F}
iq_df$ASVAB_PC1 <- pca$x[,1]
iq_df$ASVAB_PC2 <- pca$x[,2]
str(iq_df)
```

We run a linear model of log_income (response variable) vs ASVAB_PC1 and ASVAB_PC2. We control for gender, education, and race. ASVAB_PC1 and ASVAB_PC2 both have small p-values <0.001, and both have positive coefficients of about 0.06. This implies that higher PC1 and PC2 scores on average are correlated with a higher log(income).
```{r, echo=F, result=T}
fit1 <- lm(log_income ~ ASVAB_PC1 + ASVAB_PC2 + Gender + Educ + Race, data=iq_df)
summary(fit1)
```

iii. Is there any evidence to show that there is gender bias against either male or female in terms of income in the above model?

Yes, because Genderfemale is the base value, and Gendermale's beta-value is 0.67 higher than the base value. This suggests that males have higher log(income) on average than females. See the summary of the lm above.

We next build a few models for the purpose of prediction using all the information available. From now on you may use the three data sets setting (training/testing/validation) when it is appropriate. 

## 3. Trees

i. fit1: tree(Income ~ Educ + Gender, data.train) with default set up 

    a) Display the tree
    b) How many end nodes? Briefly explain how the estimation is obtained in each end nodes and deescribe the prediction equation
    c) Does it show interaction effect of Gender and Educ on Income?
    d) Predict Michelle's income

We find the tree for Educ and Gender's effect on income. There are 4 end nodes. If female, go to the left subtree. Otherwise, go to right subtree. Then in the left sub-tree, go left if Educ < 15.5; otherwise go right. The same is true for the right-subtree. This brings you to both end nodes. The prediction values in the end nodes are the average log_income values for the datapoints in data.train that fall into that partition.
The prediction equation: If female and educ < 15.5, then log_income 9.959. If female and educ > 15.5, then log_income 10.520. If male and educ < 15.5, then log_income 10.570. If male and educ > 15.5, then log_income 11.160.

There is an interactive effect of gender and education on log_income. The decision tree analysis suggests that for females, higher education (Educ > 15.5) is associated with higher income, while for males, higher education also leads to higher income, but the effect is not as pronounced as for females.

Our tree model predicts that **Michelle** has log_income 9.959322 because she is a female with educ = 13 < 15.5.

```{r, echo=F, results=T}
fit_t1 <- tree(log_income ~ Educ + Gender, data.train) 
plot(fit_t1)
title(main =  "fit_t1 tree: Educ + Gender")
text(fit_t1, pretty=0)
```
```{r, echo=F, result=T}
prediction <- predict(fit_t1, newdata = michelle)
prediction
```

ii. fit2: fit2 <- rpart(Income2005 ~., data.train, minsplit=20, cp=.009)

    a) Display the tree using plot(as.party(fit2), main="Final Tree with Rpart") 
    b) A brief summary of the fit2
    c) Compare testing errors between fit1 and fit2. Is the training error from fit2 always less than that from fit1? Is the testing error from fit2 always smaller than that from fit1? 
    d) You may prune the fit2 to get a tree with small testing error. 

fit_t2 first splits by Gender, then each subtree to the left and right is split differently. The left splits by Education and then AFQT. The right splits by Arith then Educ. There are 6 leaf nodes, each of which has the average log_income for that respective partition of the data.
```{r, echo=F, results=T}
fit_t2 <- rpart(log_income ~., data.train, minsplit=20, cp=.009) 
plot(as.party(fit_t2), main="Final Tree with Rpart")
fit_t2
```
The training and testing errors for fit1 and fit2 are printed below. Fit2 has better training error and testing error than fit1, which makes sense since fit2 has more variables, so its tree can be more precise.
```{r, echo=F, result=T}
# get training error
train.error.t1 <- sum((predict(fit_t1, data.train) - data.train$log_income)^2)
print(paste("train error t1:", train.error.t1))
train.error.t2 <- sum((predict(fit_t2, data.train) - data.train$log_income)^2)
print(paste("train error t2:", train.error.t2))

# get testing error
test.error.t1 <- sum((predict(fit_t1, data.test) - data.test$log_income)^2)
print(paste("test error t1:", test.error.t1))
test.error.t2 <- sum((predict(fit_t2, data.test) - data.test$log_income)^2)
print(paste("test error t2:", test.error.t2))
```
We predict **Michelle** using fit2 below.
```{r, echo=F, result=T}
prediction <- predict(fit_t2, newdata = michelle)
prediction
```    

With a lower cp score assigned while pruning, we cannot reach a training and testing error below what was achieved with the initial fit_t2 tree.
```{r, echo=F, result=T}
fit2.p <- prune.rpart(fit_t2, cp=0.00001)
# get training error
train.error.t2.p <- sum((predict(fit2.p, data.train) - data.train$log_income)^2)
print(paste("train error t2 pruned:", train.error.t2.p))

# get testing error
test.error.t2.p <- sum((predict(fit2.p, data.test) - data.test$log_income)^2)
print(paste("test error t2 pruned:", test.error.t2.p))
```


iii. fit3: bag two trees

    a) Take 2 bootstrap training samples and build two trees using the 
    rpart(Income2005 ~., data.train.b, minsplit=20, cp=.009). Display both trees.
    b) Explain how to get fitted values for Michelle by bagging the two trees obtained above. Do not use the predict(). 
    c) What is the testing error for the bagged tree. Is it guaranteed that the testing error by bagging the two tree always smaller that either single tree? 

We take 2 bootstrap training samples and build two trees. The trees are clearly different.
```{r, echo=F, result=T}
set.seed(1)
index1 <- sample(1808, 1808, replace=TRUE)
boot1 <- data.train[index1,]
index2 <- sample(1808, 1808, replace=TRUE)
boot2 <- data.train[index2,]

boot.1.single.full <- rpart(log_income ~., boot1, minsplit=20, cp=.009)
plot(boot.1.single.full)
title(main = "Bootstrap tree 1")
text(boot.1.single.full, pretty=0)

boot.2.single.full <- rpart(log_income ~., boot2, minsplit=20, cp=.009)
plot(boot.2.single.full)
title(main = "Bootstrap tree 2")
text(boot.2.single.full, pretty=0)
```
To find the average, we followed each tree boot1 and boot2 separately for **Michelle**'s variables and then found the value at the leaf node for each tree. We then averaged those two values. Lastly, we verified our answer by using predict to bag the two trees, and we obtained the value below:
```{r, echo=F, result=T}
# bag of two trees by averaging the two fitted equations. predict the response for michelle:
fit.bag.2.predict <- (predict(boot.1.single.full, michelle) + predict(boot.2.single.full, michelle))/2 #bagging
data.frame(fitted=fit.bag.2.predict,  obsy=michelle$log_income)
```

We find the testing error for the bagged tree. The testing errors of the boot1, boot2, and bagged trees are shown below. Our bagged error was less than that of the individual times, which was the case for several times we ran it, but not always.
```{r, echo=F, result=T}
test.error.boot1 <- sum((predict(boot.1.single.full, data.test) - data.test$log_income)^2)
print(paste("test error boot1:", test.error.boot1))
test.error.boot2 <- sum((predict(boot.2.single.full, data.test) - data.test$log_income)^2)
print(paste("test error boot2:", test.error.boot2))

test.error.bagged <- sum(((predict(boot.1.single.full, data.test) + predict(boot.2.single.full, data.test)) / 2 - data.test$log_income)^2)
print(paste("test error bagged boot1 and boot2:", test.error.bagged))
```

    
iv. fit4: Build a best possible RandomForest

    a) Show the process how you tune mtry and number of trees. Give a very high level explanation how fit4 is built.
    b) Compare the oob errors form fit4 to the testing errors using your testing data. Are you convinced that oob errors estimate testing error reasonably well.
    c) What is the predicted value for Michelle?

v. Now you have built so many predicted models (fit1 through fit4 in this section). What about build a fit5 which bags fit1 through fit4. Does fit5 have the smallest testing error?

iv. Now use `XGBoost` to build the fit6 predictive equation. Evaluate its testing error. Also briefly explain how it works. 

vii.  Summarize the results and nail down one best possible final model you will recommend to predict income. Explain briefly why this is the best choice. Finally for the first time evaluate the prediction error using the validating data set. 

viii. Use your final model to predict Michelle's income. 


    
# Problem 2: Yelp challenge 2019

**Note:** This problem is rather involved. It covers essentially all the main materials we have done so far in this semester. It could be thought as a guideline for your final project if you want when appropriate. 

Yelp has made their data available to public and launched Yelp challenge. [More information](https://www.yelp.com/dataset/). It is unlikely we will win the $5,000 prize posted but we get to use their data for free. We have done a detailed analysis in our lecture. This exercise is designed for you to get hands on the whole process. 

For this case study, we downloaded the [data](https://www.yelp.com/dataset/download) and took a 20k subset from **review.json**. *json* is another format for data. It is flexible and commonly-used for websites. Each item/subject/sample is contained in a brace *{}*. Data is stored as **key-value** pairs inside the brace. *Key* is the counterpart of column name in *csv* and *value* is the content/data. Both *key* and *value* are quoted. Each pair is separated by a comma. The following is an example of one item/subject/sample.

```{json}
{
  "key1": "value1",
  "key2": "value2"
}
```


**Data needed:** yelp_review_20k.json available in Canvas.

**yelp_review_20k.json** contains full review text data including the user_id that wrote the review and the business_id the review is written for. Here's an example of one review.

```{json}
{
    // string, 22 character unique review id
    "review_id": "zdSx_SD6obEhz9VrW9uAWA",

    // string, 22 character unique user id, maps to the user in user.json
    "user_id": "Ha3iJu77CxlrFm-vQRs_8g",

    // string, 22 character business id, maps to business in business.json
    "business_id": "tnhfDv5Il8EaGSXZGiuQGg",

    // integer, star rating
    "stars": 4,

    // string, date formatted YYYY-MM-DD
    "date": "2016-03-09",

    // string, the review itself
    "text": "Great place to hang out after work: the prices are decent, and the ambience is fun. It's a bit loud, but very lively. The staff is friendly, and the food is good. They have a good selection of drinks.",

    // integer, number of useful votes received
    "useful": 0,

    // integer, number of funny votes received
    "funny": 0,

    // integer, number of cool votes received
    "cool": 0
}
```

## Goal of the study

The goals are 

1) Try to identify important words associated with positive ratings and negative ratings. Collectively we have a sentiment analysis.  

2) To predict ratings using different methods. 

## 1. JSON data and preprocessing data

i. Load *json* data

The *json* data provided is formatted as newline delimited JSON (ndjson). It is relatively new and useful for streaming.
```{json}
{
  "key1": "value1",
  "key2": "value2"
}
{
  "key1": "value1",
  "key2": "value2"
}
```

The traditional JSON format is as follows.
```{json}
[{
  "key1": "value1",
  "key2": "value2"
},
{
  "key1": "value1",
  "key2": "value2"
}]
```


We use `stream_in()` in the `jsonlite` package to load the JSON data (of ndjson format) as `data.frame`. (For the traditional JSON file, use `fromJSON()` function.)

```{r}
pacman::p_load(jsonlite)
yelp_data <- jsonlite::stream_in(file("data/yelp_review_20k.json"), verbose = F)
str(yelp_data)  
# different JSON format
# tmp_json <- toJSON(yelp_data[1:10,])
# fromJSON(tmp_json)
```

**Write a brief summary about the data:**

a) Which time period were the reviews collected in this data?

b) Are ratings (with 5 levels) related to month of the year or days of the week? Only address this through EDA please. 


ii. Document term matrix (dtm)
 
 Extract document term matrix for texts to keep words appearing at least .5% of the time among all 20000 documents. Go through the similar process of cleansing as we did in the lecture. 

a) Briefly explain what does this matrix record? What is the cell number at row 100 and column 405? What does it represent?

b) What is the sparsity of the dtm obtained here? What does that mean?

iii. Set the stars as a two category response variable called rating to be “1” = 5,4 and “0”= 1,2,3. Combine the variable rating with the dtm as a data frame called data2. 

## Analysis

Get a training data with 13000 reviews and the 5000 reserved as the testing data. Keep the rest (2000) as our validation data set. 

## 2. LASSO

i. Use the training data to get Lasso fit. Choose lambda.1se. Keep the result here.

ii. Feed the output from Lasso above, get a logistic regression. 
	
a) Pull out all the positive coefficients and the corresponding words. Rank the coefficients in a decreasing order. Report the leading 2 words and the coefficients. Describe briefly the interpretation for those two coefficients. 

b) Make a word cloud with the top 100 positive words according to their coefficients. Interpret the cloud briefly.

c) Repeat i) and ii) for the bag of negative words.

d) Summarize the findings. 

iii. Using majority votes find the testing errors
	i) From Lasso fit in 3)
	ii) From logistic regression in 4)
	iii) Which one is smaller?

## 3. Random Forest  

i. Briefly summarize the method of Random Forest

ii. Now train the data using the training data set by RF. Get the testing error of majority vote. Also explain how you tune the tuning parameters (`mtry` and `ntree`). 

## 4. Boosting 

Now use `XGBoost` to build the fourth predictive equation. Evaluate its testing error. 

## 5. Ensemble model

i. Take average of some of the  models built above (also try all of them) and this gives us the fifth model. Report it's testing error. (Do you have more models to be bagged, try it.)


## 6. Final model

Which classifier(s) seem to produce the least testing error? Are you surprised? Report the final model and accompany the validation error. Once again this is THE only time you use the validation data set.  For the purpose of prediction, comment on how would you predict a rating if you are given a review (not a tm output) using our final model? 











